Weighted Model Counting with Conditional Measures

### 2020-07-26 paper.tex
Same

\todo[inline]{
  Notation: make up my mind about $a, b$ vs. $x, y$ and stick to it (maybe $x,
  y$?).
}

@@

\begin{remark}
  For any function $A\colon 2^X \to \mathbb{R}_{\ge 0}$, $A + \overline{A} = 1$.
\end{remark}

@@

\todo[inline]{Describe \cref{tbl:notation_example} as an example.}
\todo[inline]{Feedback: explain with examples: models $=$ atoms of algebra,
  formulas $=$ all elements.}

\begin{definition} \label{def:wmc}
  Let $U$ be a set.
  \begin{itemize}
  \item A \emph{measure} is a function $M\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$
    such that $M(\bot) = 0$ and
    \[
      M(a \lor b) = M(a) + M(b)
    \]
    for all $a, b \in 2^{2^U}$ whenever $a \land b = \bot$.
  \item A \emph{weight function} is a function $W\colon 2^U \to \mathbb{R}_{\ge
      0}$.
  \item A weight function is \emph{factored} if $W = \prod_{x \in U} W_x$ for
    some functions $W_x\colon 2^{\{x\}} \to \mathbb{R}_{\ge 0}$.
  \item A measure $M\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ is
    \emph{factorable} if there exists a factored weight function $W\colon 2^U
    \to \mathbb{R}_{\ge 0}$ that induces $M$.
  \end{itemize}
\end{definition}

\begin{proposition} \label{prop:measure}
  Every weight function induces a measure, i.e, for any weight function $W\colon
  2^U \to \mathbb{R}_{\ge 0}$, the function $M_W\colon 2^{2^U} \to
  \mathbb{R}_{\ge 0}$, defined as
  \[
    M_W(x) = \begin{cases}
      0 & \text{if } x = \bot \\
      W(u) & \text{if } x = \{ u \} \\
      \sum_{\{u\} \le x} W(u) & \text{otherwise,}
    \end{cases}
  \]
  is a measure.
\end{proposition}
\begin{proof}
  \todo[inline]{Prove that $M_W$ satisfies the definition of a measure.}
\end{proof}

The process of calculating the value of $M_W(x)$ for some $x \in 2^{2^U}$ with a
given definition of $W$ is known as \emph{weighted model counting}.

\begin{table}
  \caption{(in the same order)}
  \label{tbl:notation_example}
  \centering
  \begin{tabular}{lcc}
    \toprule
    & Set-theoretic notation & Boolean-algebraic notation \\
    \midrule
    Atoms (elements of $U$) & $a, b$ & $a, b$ \\
    \rowcolor{gray!10} Models (elements of $2^U$) & $\emptyset, \{a\}, \{b\}, \{a, b\}$ & $\neg a \land \neg b, a \land \neg b, \neg a \land b, a \land b$ \\
    & $\{ \emptyset, \{a\}, \{b\}, \{a, b\} \}$ & $\top$ \\
    & $\{ \emptyset, \{a\}, \{b\} \}, \{ \emptyset, \{a\}, \{a, b\} \}$ & $\neg a \lor \neg b, a \to b$ \\
    & $\{ \emptyset, \{b\}, \{a, b\} \}, \{ \{a\}, \{b\}, \{a, b\} \}$ & $b \to a, a \lor b$ \\
    & $\{\emptyset, \{a\}\}, \{\emptyset, \{b\}\}, \{\emptyset, \{a, b\}\}$ & $\neg b, \neg a, a \leftrightarrow b$ \\
    & $\{\{a\}, \{b\}\}, \{\{a\}, \{a, b\}\}, \{\{b\}, \{a, b\}\}$ & $(a \land \neg b) \lor (b \land \neg a), a, b$ \\
    & $\{\emptyset\}, \{\{a\}\}, \{\{b\}\}, \{\{a, b\}\}$ & $\neg a \land \neg b, a \land \neg b, \neg a \land b, a \land b$ \\
    \multirow{-7}{*}{Formulas (elements of $2^{2^U}$)} & $\emptyset$ & $\bot$ \\
    \bottomrule
  \end{tabular}
\end{table}

\begin{example}[Relation to the classical view of WMC] \label{example:construction}
  Let $\mathcal{L}$ be a propositional logic with a set of atoms $U = \{ p, q
  \}$. Let $w : \{ p, q, \neg p, \neg q \} \to \mathbb{R}_{\ge 0}$ be the
  \emph{weight function} defined by
  \begin{align*}
    w(p) = 0.3, \quad w(\neg p) = 0.7, \quad w(q) = 0.2, \quad w(\neg q) = 0.8.
  \end{align*}
  Let $\Delta$ be a theory in $\mathcal{L}$ with a sole axiom $p$. Then
  $\Delta$ has two models, i.e., $\{ p, q \}$ and $\{ p, \neg q \}$. The
  \emph{weighted model count} (WMC) \cite{DBLP:journals/ai/ChaviraD08} of
  $\Delta$ is then
  \begin{equation} \label{eq:wmc_example}
    \mathrm{WMC}(\Delta) = \sum_{\omega \models \Delta} \prod_{\omega \models l} w(l) = w(p)w(q) + w(p)w(\neg q) = 0.3.
  \end{equation}

  Alternatively, we can define $W_p\colon 2^{\{p\}} \to \mathbb{R}_{\ge 0}$ as
  \[
    W_p(\{ p \}) = 0.3, \quad W_p(\emptyset) = 0.7
  \]
  and $W_q\colon 2^{\{q\}} \to \mathbb{R}_{\ge 0}$ as
  \[
    W_q(\{ q \}) = 0.2, \quad W_q(\emptyset) = 0.8.
  \]
  Let $M$ be the measure on $2^{2^U}$ induced by $W = W_p \cdot W_q$. Then,
  equivalently to \cref{eq:wmc_example}, we can write
  \[
    M(p) = W(\{ p, q \}) + W(\{ p \}) = W_p(\{p\})W_q(\{q\}) +
    W_p(\{p\})W_q(\emptyset) = 0.3.
  \]

  Given the theory $\Delta$, we can compute the probability of a query $q$ as
  \cite{DBLP:conf/uai/Belle17,DBLP:conf/aaai/SangBK05}
  \[
    \Pr_{\Delta, w}(q) = \frac{\WMC(\Delta \land q)}{\WMC(\Delta)}.
  \]
  The same thing can be accomplished using the algebraic formulation, with $M$
  replacing $\mathrm{WMC}$.
\end{example}

For the rest of the paper, we let $U$ be the set of (logical) atoms. We use
set-theoretic notation for $2^U$ and Boolean-algebraic notation for $2^{2^U}$,
except for (Boolean) atoms in $2^{2^U}$ that are denoted as $\{x\}$ for some
model $x \in 2^U$.

@@

\begin{itemize}
\item[F] Give a concrete example of something impossible to represent using WMC.
\item[F] Can you say something here about factorized vs non-factorized
  weight function definitions? That is, factorized is when $w$ maps literals to
  $R_{\ge 0}$, non-factorized is when $w$ maps models to $R_{\ge 0}$ and
  \begin{itemize}
  \item come up with nice example when non-factorized weights are intuitive;
  \item clarify that the factorized definition have is w.r.t. models, in case some
    one gets confused. [It doesn't have to be, if the BA is not free---P.]
  \end{itemize}
\end{itemize}

@@

\subsection{Bayesian Network Encodings}

\begin{itemize}
\item Encodings:
  \begin{itemize}
  \item \texttt{d02} \cite{DBLP:conf/kr/Darwiche02}
  \item \texttt{sbk05} \cite{DBLP:conf/aaai/SangBK05}
  \item \texttt{cd05} \cite{DBLP:conf/ijcai/ChaviraD05}
  \item \texttt{cd06} \cite{DBLP:conf/sat/ChaviraD06} (supposed to be the best)
  \item \texttt{db20} (mine)
  \end{itemize}
\item \texttt{cd05} relaxes the encoding so much that extra models become
  possible. They are supposed to be filtered out by the algorithm, but mine
  can't do that because it doesn't deal with models. Same for \texttt{cd06}
  because it's based on \texttt{cd05}.
\end{itemize}

@@

Cover:
\begin{itemize}
\item ADDs provide an efficient way to manipulate functions from BAs/power sets
  \cite{DBLP:journals/fmsd/BaharFGHMPS97}.
\item background reading
  \begin{itemize}
  \item Compiling Bayesian Networks Using Variable Elimination (Chavira and
    Darwiche) \cite{DBLP:conf/ijcai/ChaviraD07}
  \item On the Relationship between Sum-Product Networks and Bayesian Networks
    (Zhao et al.) \cite{DBLP:conf/icml/ZhaoMP15}
  \end{itemize}
\end{itemize}

@@

@@

\section{Experimental Comparison}

\begin{itemize}
\item We don't compare 'compile times' because our encoding time is linear, so
  we would easily beat everyone else.
\item When the Bayesian network has an evidence file, we compute the probability
  of evidence. Otherwise, let $X$ denote the last-mentioned node in the Bayesian
  network. If $\mathsf{true}$ is a valid value of $X$, we compute the marginal
  probability of $X = \mathsf{true}$. Otherwise, we pick the first value of $X$
  and calculate its marginal probability. This applies to the Grid data set (as
  intended) and also to two instances of Plan Reconstruction and roughly half of
  the instances from 2004-PGM that have empty evidence files.
\item After the experiments are finished, note the processor, memory per thread,
  and add the following acknowledgment.
\item All other encodings are implemented in
  Ace~3.0\footnote{\url{http://reasoning.cs.ucla.edu/ace/}} and should be
  compiled with \texttt{-encodeOnly} (i.e., don't compile the CNF into an AC)
  and \texttt{-noEclause} (i.e., only use standard syntax) flags.
\item Datasets
  \begin{itemize}
  \item binary Bayesian networks from Sang et
    al.\footnote{\url{https://www.cs.rochester.edu/u/kautz/Cachet/}}
    \cite{DBLP:conf/aaai/SangBK05}
    \begin{itemize}
    \item Grid (networks) (ratio 75 means that 75\% of the nodes are
      deterministic),
    \item Plan recognition (problems),
    \item Deterministic quick medical reference (what do the numbers mean? the
      README doesn't say).
    \end{itemize}
  \item Bayesian networks available with Ace
    \begin{itemize}
    \item \texttt{2004-pgm} \cite{DBLP:journals/ijar/ChaviraDJ06} (binary)
    \item \texttt{2005-ijcai} \cite{DBLP:conf/ijcai/ChaviraD05}. The Genie/Smile
      files have their own citation data that I should probably extract. This is
      the only dataset that has some non-binary networks.
    \item \texttt{2006-ijar} \cite{DBLP:journals/ijar/ChaviraDJ06} (binary)
    \end{itemize}
  \end{itemize}
\end{itemize}

>>>>

@@

@@

\begin{definition}[Measures and weight functions] \label{def:wmc}
  Let $U$ be a set.
  \begin{itemize}
  \item A \emph{measure} is a function $M\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$
    such that $M(\bot) = 0$ and
    \[
      M(a \lor b) = M(a) + M(b)
    \]
    for all $a, b \in 2^{2^U}$ whenever $a \land b = \bot$.
  \item A \emph{weight function} is a function $W\colon 2^U \to \mathbb{R}_{\ge
      0}$.
  \item A weight function is \emph{factored} if $W = \prod_{x \in U} W_x$ for
    some functions $W_x\colon 2^{\{x\}} \to \mathbb{R}_{\ge 0}$, $x \in U$.
  \item We say that a weight function $W\colon 2^U \to \mathbb{R}_{\ge 0}$
    \emph{induces} a measure $M_W\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ if
    \begin{equation} \label{eq:induced_measure}
      M_W(x) = \sum_{\{u\} \le x} W(u).
    \end{equation}
  \item A measure $M\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ is
    \emph{factorable} if there exists a factored weight function $W\colon 2^U
    \to \mathbb{R}_{\ge 0}$ that induces $M$.
  \end{itemize}
\end{definition}

\begin{lemma} \label{prop:measure}
  The function $M_W$, as defined by \cref{eq:induced_measure}, is a measure.
\end{lemma}
\begin{proof}
  Note that $M_W(\bot) = 0$ since there are no atoms below $\bot$. Let $a, b \in
  2^{2^{U}}$ be such that $a \land b = \bot$. By elementary properties of
  Boolean algebras, all atoms below $a \lor b$ are either below $a$ or below
  $b$. Moreover, none of them can be below both $a$ and $b$ because then they
  would have to be below $a \land b = \bot$. Thus
  \[
    M_W(a \lor b) = \sum_{\{u\} \le a \lor b} W(u) = \sum_{\{u\} \le a} W(u) +
    \sum_{\{u\} \le b} = M_W(a) + M_W(b)
  \]
  as required.
\end{proof}

In this formulation, the process of calculating the value of $M_W(x)$ for some
$x \in 2^{2^U}$ with a given definition of $W$ is known as \emph{weighted model
  counting}.

\subsection{Relation to the Classical (Logic-Based) View of WMC}

\todo[inline]{Would it be more accurate to replace the word `model' with
  `interpretation' in most situations?}

\begin{table}
  \caption{An overview of notation for a logic defined over two atoms. The
    elements in both columns are listed in the same order.}
  \label{tbl:notation_example}
  \centering
  \begin{tabular}{lcc}
    \toprule
    Name in logic & Set-theoretic notation & Boolean-algebraic notation \\
    \midrule
    Atoms (elements of $U$) & $a, b$ & $a, b$ \\
    \rowcolor{gray!10} Models (elements of $2^U$) & $\emptyset, \{a\}, \{b\}, \{a, b\}$ & $\neg a \land \neg b, a \land \neg b, \neg a \land b, a \land b$ \\
    & $\{ \emptyset, \{a\}, \{b\}, \{a, b\} \}$ & $\top$ \\
    & $\{ \emptyset, \{a\}, \{b\} \}, \{ \emptyset, \{a\}, \{a, b\} \}$ & $\neg a \lor \neg b, a \to b$ \\
    & $\{ \emptyset, \{b\}, \{a, b\} \}, \{ \{a\}, \{b\}, \{a, b\} \}$ & $b \to a, a \lor b$ \\
    & $\{\emptyset, \{a\}\}, \{\emptyset, \{b\}\}, \{\emptyset, \{a, b\}\}$ & $\neg b, \neg a, a \leftrightarrow b$ \\
    & $\{\{a\}, \{b\}\}, \{\{a\}, \{a, b\}\}, \{\{b\}, \{a, b\}\}$ & $(a \land \neg b) \lor (b \land \neg a), a, b$ \\
    & $\{\emptyset\}, \{\{a\}\}, \{\{b\}\}, \{\{a, b\}\}$ & $\neg a \land \neg b, a \land \neg b, \neg a \land b, a \land b$ \\
    \multirow{-7}{*}{Formulas (elements of $2^{2^U}$)} & $\emptyset$ & $\bot$ \\
    \bottomrule
  \end{tabular}
\end{table}

Let $\mathcal{L}$ be a propositional logic with atoms $U = \{ a, b \}$. See
\cref{tbl:notation_example} for an overview of the models and formulas in
$\mathcal{L}$ as well as their set-theoretic and Boolean-algebraic
representations. Note the differences in the meaning of the word `atom' between
the logical and the Boolean-theoretic interpretations. In the Boolean algebra
$2^{2^U}$, an atom is a set with a single element which corresponds to a model
of $\mathcal{L}$ (also known as an element of $2^U$); whereas an atom of
$\mathcal{L}$ is an atomic formula, i.e., an element of $U$. We will primarily
use the term `atom' to mean the former.

Let $w\colon \{ a, b, \neg a, \neg b \} \to \mathbb{R}_{\ge 0}$
be the \emph{weight function} defined by
\begin{align*}
  w(a) = 0.3, \quad w(\neg a) = 0.7, \quad w(b) = 0.2, \quad w(\neg b) = 0.8.
\end{align*}
Let $\Delta$ be a theory in $\mathcal{L}$ with a sole axiom $a$. Then
$\Delta$ has two models, i.e., $\{ a, b \}$ and $\{ a, \neg b \}$. The
\emph{weighted model count} (WMC) \cite{DBLP:journals/ai/ChaviraD08} of
$\Delta$ is then
\begin{equation} \label{eq:wmc_example}
  \mathrm{WMC}(\Delta) = \sum_{\omega \models \Delta} \prod_{\omega \models l} w(l) = w(a)w(b) + w(a)w(\neg b) = 0.3.
\end{equation}

Alternatively, we can define $W_a\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ as
\[
  W_a(\{ a \}) = 0.3, \quad W_a(\emptyset) = 0.7
\]
and $W_b\colon 2^{\{q\}} \to \mathbb{R}_{\ge 0}$ as
\[
  W_b(\{ b \}) = 0.2, \quad W_b(\emptyset) = 0.8.
\]
Let $M$ be the measure on $2^{2^U}$ induced by $W = W_a \cdot W_b$. Then,
equivalently to \cref{eq:wmc_example}, we can write
\[
  M(a) = W(\{ a, b \}) + W(\{ a \}) = W_a(\{a\})W_b(\{b\}) +
  W_a(\{a\})W_b(\emptyset) = 0.3.
\]

Given the theory $\Delta$, we can also compute the probability of a query $b$ as
\cite{DBLP:conf/uai/Belle17,DBLP:conf/aaai/SangBK05}
\[
  \Pr_{\Delta, w}(b) = \frac{\WMC(\Delta \land b)}{\WMC(\Delta)}.
\]
The same thing can be accomplished using the algebraic formulation, with $M$
replacing $\mathrm{WMC}$.

In the rest of the paper, for any set $U$, we will use set-theoretic notation
for $2^U$ and Boolean-algebraic notation for $2^{2^U}$, except for (Boolean)
atoms in $2^{2^U}$ that are denoted as $\{x\}$ for some model $x \in 2^U$.

@@

\begin{itemize}
\item[F] Give a concrete example of something impossible to represent using WMC.
\item[F] Can you say something here about factorized vs non-factorized
  weight function definitions? That is, factorized is when $w$ maps literals to
  $R_{\ge 0}$, non-factorized is when $w$ maps models to $R_{\ge 0}$ and
  \begin{itemize}
  \item come up with nice example when non-factorized weights are intuitive;
  \end{itemize}
\end{itemize}

@@

\subsection{Using WMC to Perform Inference on Bayesian Networks}

\todo[inline]{Provide a reference (and an explanation) for d-DNNF.}
\todo[inline]{If I'm citing Cachet, then maybe I should also mention what
  compilation tools they're using in their experiments.}

Hitherto, four techniques have been proposed for encoding Bayesian networks into
instances of WMC. We will identify them based on the initials of authors as well
as publications years: \texttt{d02} \cite{DBLP:conf/kr/Darwiche02},
\texttt{sbk05} \cite{DBLP:conf/aaai/SangBK05}, \texttt{cd05}
\cite{DBLP:conf/ijcai/ChaviraD05}, and \texttt{cd06}
\cite{DBLP:conf/sat/ChaviraD06}. Below we friendly summarise the observed
performance differences among them. Sang et al. \cite{DBLP:conf/aaai/SangBK05}
claim that \texttt{sbk05} is a smaller encoding than \texttt{d02} with respect
to both the number of clauses and the number of variables but provide no
experimental comparison. Chavira and Darwiche \cite{DBLP:conf/ijcai/ChaviraD05}
compare \texttt{cd05} with \texttt{d02} by measuring the time it takes to
compile either encoding into an arithmetic circuit (but do not measure inference
time). The more recent encoding \texttt{cd05} always compiles faster and results
in a smaller arithmetic circuit (as measured by the number of edges). In their
subsequent paper, the same authors perform two sets of experiments (that are
relevant to this summary) \cite{DBLP:conf/sat/ChaviraD06}. First, they compile
\texttt{cd05} and \texttt{cd06} encodings into d-DNNF, measuring both
compilation time and numbers of edges in the d-DNNF diagram. The results are
mostly in favour of \texttt{cd06}. Second, they compare the inference time of
\texttt{sbk05} run with Cachet \cite{DBLP:conf/sat/SangBBKP04} with the compile
times of \texttt{cd05} and \texttt{cd06}, but only on five (types of) instances.
In these experiments, \texttt{cd06} is always faster than \texttt{cd05}, while
the comparison with \texttt{sbk05} is mixed. Sometimes \texttt{cd06} is orders
of magnitude faster than \texttt{sbk05}, sometimes slightly slower. The
performance difference between \texttt{sbk05} and \texttt{cd05} is even harder
to judge: \texttt{sbk05} is better on three out of five instances and worse on
the remaining two. Based on this description, one would expect \texttt{cd06} to
be faster than both \texttt{cd05} and \texttt{sbk05}, both of which should be
faster than \texttt{d02}. The experiments in \cref{sec:experiments}, however,
strongly disagree with this prediction, showing that the quality of an encoding
depends strongly on the underlying search algorithm or compilation technique.

@@

Cover:
\begin{itemize}
\item ADDs provide an efficient way to manipulate functions from BAs/power sets
  \cite{DBLP:journals/fmsd/BaharFGHMPS97}.
\item background reading
  \begin{itemize}
  \item Compiling Bayesian Networks Using Variable Elimination (Chavira and
    Darwiche) \cite{DBLP:conf/ijcai/ChaviraD07}
  \item On the Relationship between Sum-Product Networks and Bayesian Networks
    \cite{DBLP:conf/icml/ZhaoMP15}
  \item Using ROBDDs for Inference in Bayesian Networks with Troubleshooting as
    an Example \cite{DBLP:conf/uai/NielsenWJK00}
  \item SPUDD: Stochastic Planning using Decision Diagrams
    \cite{DBLP:conf/uai/HoeySHB99}
  \end{itemize}
\end{itemize}

@@

\todo[inline]{Even though a lot of papers don't do this, I want a reference and
  a bit of a (historical/contextual) description here.}

@@

\section{Experimental Comparison} \label{sec:experiments}

\begin{itemize}
\item We don't compare 'compile times' because our encoding time is linear, so
  we would easily beat everyone else.
\item When the Bayesian network has an evidence file, we compute the probability
  of evidence. Otherwise, let $X$ denote the last-mentioned node in the Bayesian
  network. If $\mathsf{true}$ is a valid value of $X$, we compute the marginal
  probability of $X = \mathsf{true}$. Otherwise, we pick the first value of $X$
  and calculate its marginal probability. This applies to the Grid data set (as
  intended) and also to two instances of Plan Reconstruction and roughly half of
  the instances from 2004-PGM that have empty evidence files.
\item After the experiments are finished, note the processor, memory per thread,
  and add the following acknowledgment.
\item All other encodings are implemented in
  Ace~3.0\footnote{\url{http://reasoning.cs.ucla.edu/ace/}} and should be
  compiled with \texttt{-encodeOnly} (i.e., don't compile the CNF into an AC)
  and \texttt{-noEclause} (i.e., only use standard syntax) flags.
\item Datasets
  \begin{itemize}
  \item binary Bayesian networks from Sang et
    al.\footnote{\url{https://www.cs.rochester.edu/u/kautz/Cachet/}}
    \cite{DBLP:conf/aaai/SangBK05}
    \begin{itemize}
    \item Grid (networks) (ratio 75 means that 75\% of the nodes are
      deterministic),
    \item Plan recognition (problems),
    \item Deterministic quick medical reference (what do the numbers mean? the
      README doesn't say).
    \end{itemize}
  \item Bayesian networks available with Ace
    \begin{itemize}
    \item \texttt{2004-pgm} \cite{DBLP:journals/ijar/ChaviraDJ06} (binary)
    \item \texttt{2005-ijcai} \cite{DBLP:conf/ijcai/ChaviraD05}. The Genie/Smile
      files have their own citation data that I should probably extract. This is
      the only dataset that has some non-binary networks.
    \item \texttt{2006-ijar} \cite{DBLP:journals/ijar/ChaviraDJ06} (binary)
    \end{itemize}
  \end{itemize}
\item \texttt{cd05} relaxes the encoding so much that extra models become
  possible. They are supposed to be filtered out by the algorithm, but mine
  can't do that because it doesn't deal with models. Same for \texttt{cd06}
  because it's based on \texttt{cd05}.
\end{itemize}
